---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# phenologysim

`phenologysim` is an R package for modelling temperature-driven phenology and life-cycle development.

The package provides:

- Temperature-dependent development functions
- Developmental variability functions
- Chilling accumulation models
- Cohort-based life-stage simulation
- Daylength-dependent life stages
- Population simulation tools
- Visualisation functions
- Example datasets and models

The framework is designed for insects, plants, and other organisms whose development depends on environmental conditions.

## Installation

Install the development version from GitHub:

```{r eval = FALSE}
install.packages("remotes")

remotes::install_github("John-Kean/phenologysim")
```

Load the package:

```{r eval = FALSE}
library(phenologysim)
```

## Quick start

Create some example environmental drivers:

```{r eval = FALSE}
drivers <- example_varying_drivers()
```

Run a population simulation:

```{r eval = FALSE}
results <- run_simulation(
  population = example_lifestages,
  initialise = c(eggs = 1000),
  timestep = example_timestep,
  drivers = drivers
)
```

Plot the population trajectory:

```{r eval = FALSE}
plot_population_trajectory(
  results[, c("date", "eggs", "larvae", "pupae", "adults")]
)
```

## Development models

The package includes a range of commonly-used phenology models:

### Linear models

- `calendar_development()`
- `linear_development()`

### Exponential models

- `exponential_development()`
- `expbase_development()`

### Sigmoid models

- `sigmoid_development()`

### Gaussian models

- `Gaussian_development()`

### Logan models

- `Logan1_development()`
- `Logan2_development()`

### Lactin models

- `Lactin1_development()`
- `Lactin2_development()`

### Analytis models

- `Analytis1_development()`
- `Analytis2_development()`
- `Analytis3_development()`
- `AnalytisAllahyari_development()`

### Briere models

- `Briere1_development()`
- `Briere2_development()`

### Beta model

- `Beta_development()`

## Variability models

To represent variation among individuals within a population:

- `step_variability()`
- `logistic_variability()`
- `Regniere1_variability()`
- `Regniere2_variability()`
- `Weibull_variability()`
- `cum_normal_variability()`

## Life-stage modelling

The package provides object-oriented S3 classes for modelling populations.

### Standard life stage

```{r eval = FALSE}
larvae <- new_lifestage(
  name = "Larvae",
  devFunction = linear_development,
  devParameters = list(b = 10, q = 200),
  varFunction = cum_normal_variability,
  varParameters = list(sd = 0.1)
)
```

### Chilled life stage

```{r eval = FALSE}
eggs <- new_chilled_lifestage(
  name = "Eggs",
  devFunction = linear_development,
  devParameters = list(b = 7.4, q = 428),
  varFunction = cum_normal_variability,
  varParameters = list(sd = 0.1),
  chillFunction = linear_chill,
  chillParameters = list(T_chill = 10),
  chillResponse = function(chill) {
    1 - 0.46 * exp(-0.0005 * chill)
  }
)
```

### Daylength-dependent life stage

```{r eval = FALSE}
diapause <- new_daylength_lifestage(
  name = "Diapause",
  Threshold = 13.5,
  ReqTrend = 1,
  Variation = 0.5
)
```

## Environmental drivers

Prepare environmental data for simulations:

```{r eval = FALSE}
driver_data <- prepare_drivers(weather_data)
```

Required columns include:

| Column | Description |
|----------|-----------|
| date | Date |
| Tmean | Mean temperature (°C) |
| Tmin | Minimum temperature (°C) |
| Tmax | Maximum temperature (°C) |
| daylength | Daylength (hours) |

If daylength is not supplied, it can be calculated from latitude.

## Example simulation

Run the built-in example model:

```{r eval = FALSE}
example_phenology_simulation()
```

Or simulate multiple years:

```{r eval = FALSE}
example_phenology_simulation(length = 730)
```

## Plotting

Visualise drivers:

```{r eval = FALSE}
plot_drivers(example_varying_drivers())
```

Visualise population trajectories:

```{r eval = FALSE}
plot_population_trajectory(results)
```

## References

Many development functions are based on published phenology and insect development models, including:

- Logan et al. (1976)
- Régnière (1984)
- Casagrande et al. (1987)
- Logan (1988)
- Brière et al. (1999)
- Han et al. (2000)
- Walgama & Zalucki (2006)
- Mirhosseini et al. (2017)

See individual function documentation for full citations.

## Author

**John Kean**  
Bioeconomy Science Institute, New Zealand

## License

GPL-3

