---
title: "Introduction to phenologysim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phenologysim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

# Introduction

`phenologysim` is an R package for modelling phenology and population dynamics in organisms whose development depends on environmental conditions.

```{r setup}

# remotes::install_github("John-Kean/phenologysim")

library(phenologysim)

```

The package provides:

- Temperature-dependent development models
- Developmental variability models
- Chilling-dependent development
- Daylength-dependent life stages
- Cohort-based population simulation
- Visualisation tools
- Example models and datasets

Although originally developed for insect phenology modelling, the framework can be applied to any organism where progression through life stages depends on temperature, time, chilling accumulation or photoperiod.


# Core concepts

A phenology simulation in `phenologysim` consists of four main components:

1. **Environmental drivers** such as temperature and daylength.
2. **Life stages** such as eggs, larvae, pupae and adults.
3. **A timestep function** describing transitions between life stages.
4. **A simulation engine** that applies the timestep repeatedly through time.

The package includes helper functions for each of these components.


# Environmental drivers

Environmental conditions are provided as a driver data frame containing:

- date
- dayfraction
- degreesC
- daylength
- daytrend

If daily min/max/mean temperatures are provided, these can be converted to the required format using `prepare_drivers()`.

```{r create-drivers}

# Specify daily min/max/mean temperatures
cos <- 5 * cos(seq(0, 2 * pi, length.out = 365))
drivers_raw <- data.frame(
  date = seq.Date(as.Date("2020-07-01"), by = "day", length.out = 365),
  Tmin  = rnorm(365, mean = 10 - cos, sd = 1),
  Tmax  = rnorm(365, mean = 20 - cos, sd = 1),
  Tmean = rnorm(365, mean = 15 - cos, sd = 1)
) |>
  dplyr::mutate(daylength = geosphere::daylength(lat = -37.44, doy = lubridate::yday(date)))

# Convert to required format
drivers <- prepare_drivers(drivers_raw)
head(drivers)

```

The generated data contain seasonal variation in both temperature and daylength.

```{r plot-drivers}

plot_drivers(drivers)

```


# Life stages

Life stages are represented as S3 objects.

A typical stage consists of:

- a development model
- development parameters
- a variability model
- variability parameters

For example:

```{r create-larvae}

# Create lifestage for larvae
larvae <- new_lifestage(
  name = "Larvae",
  devFunction = linear_development,
  devParameters = list(
    b = 10,
    q = 200
  ),
  varFunction = cum_normal_variability,
  varParameters = list(
    sd = 0.1
  )
)

larvae

```

A life stage may contain multiple cohorts of different ages and developmental states.

## Development models

Many development models are available.

For example, a simple linear degree-day model:

```{r fig-linear-development}

curve(
  linear_development(x, b = 10, q = 200),
  from = 0,
  to = 35,
  xlab = "Temperature (°C)",
  ylab = "Development rate (/d)"
)

```

More complex nonlinear models are also available, including:

- beta_development()
- briere1_development()
- briere2_development()
- gaussian_development()
- lactin1_development()
- lactin2_development()
- logan1_development()
- logan2_development()
- sigmoid_development()


## Variability models

Development is rarely identical among individuals.

Variability functions convert accumulated development into the proportion of a population that has completed a stage.

For example:

```{r fig-cum-normal-variability}

curve(
  cum_normal_variability(x, sd = 0.1),
  from = 0.5,
  to = 1.5,
  xlab = "Relative development",
  ylab = "Proportion of population completed"
)

```

Alternative variability functions include:

- logistic_variability()
- regniere1_variability()
- regniere2_variability()
- step_variability()
- weibull_variability()


A whole set of life stages can be set up as a named list.

```{r example-life-stages}


eggs <- new_chilled_lifestage(
  name = "Eggs",
  devFunction = linear_development,
  devParameters = list(b = 7.4, q = 428),
  varFunction = cum_normal_variability,
  varParameters = list(sd = 0.1),
  chillFunction = linear_chill,
  chillParameters = list(T_chill = 10),
  chillResponse = function(cumChill) { 1 - 0.46 * exp(-0.0005 * cumChill) }
)



example_lifestages <- list(

  eggs = new_chilled_lifestage(
    name = "Eggs",
    devFunction = linear_development,
    devParameters = list(b = 7.4, q = 428),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1),
    chillFunction = linear_chill,
    chillParameters = list(T_chill = 10),
    chillResponse = function(cumChill) { 1 - 0.46 * exp(-0.0005 * cumChill) }
  ),

  larvae = new_lifestage(
    name = "Larvae",
    devFunction = briere2_development,
    devParameters = list(
      T0 = 12,
      Tu = 33,
      a = 4.7e-5,
      b = 4.2
    ),
    varFunction = weibull_variability,
    varParameters = list(a = 0.8, c = 3)
  ),

  pupae = new_lifestage(
    name = "Pupae",
    devFunction = linear_development,
    devParameters = list(b = 10, q = 200),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1)
  ),

  adults = new_lifestage(
    name = "Adults",
    devFunction = calendar_development,
    devParameters = list(Q = 14),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1)
  )
)

```


# Population processes

Authors can 

```{r}

example_timestep <- function(
  population,
  drivers,
  survival = c(
    eggs   = 0.999,
    larvae = 0.99,
    pupae  = 0.99,
    adults = 0.9
  ),
  fecundity = 0.2) {

  p <- population
  d <- drivers

  # Survival and development
  p$eggs   <- p$eggs   |> survive(d, survival["eggs"])   |> develop(d)
  p$larvae <- p$larvae |> survive(d, survival["larvae"]) |> develop(d)
  p$pupae  <- p$pupae  |> survive(d, survival["pupae"])  |> develop(d)
  p$adults <- p$adults |> survive(d, survival["adults"]) |> develop(d)

  # Stages mature
  p$eggs   <- p$eggs   |> add_cohort(rpois(1, total(p$adults) * fecundity))
  p$larvae <- p$larvae |> add_cohort(maturing(p$eggs))
  p$pupae  <- p$pupae  |> add_cohort(maturing(p$larvae))
  p$adults <- p$adults |> add_cohort(maturing(p$pupae))

  p
}


```


# Generic simulation engine

The package offers an optional generic simulation engine, `run_simulation()`. The user supplies:

- a population definition
- initial population sizes
- a timestep function
- environmental drivers

The bundled example model implements a complete insect life cycle.


# Running the example simulation

```{r run-simulation}

set.seed(123)

# results <- run_simulation(
#   population = example_lifestages,
#   initialise = c(eggs = 1000),
#   timestep = example_timestep,
#   drivers = drivers
# )
# 
# head(results)

```

The output contains both environmental variables and life-stage abundances through time.

## Visualising population dynamics

The trajectory of each life stage can be visualised using:

```{r, fig.height=5}

# plot_population_trajectory(
#   results[, c("date", "eggs", "larvae", "pupae", "adults")]
# )

```

This shows the seasonal dynamics generated by the example model.

A line plot may also be used:

```{r, eval = FALSE}

# plot_population_trajectory(
#   results[, c("date", "eggs", "larvae", "pupae", "adults")],
#   type = "line"
# )

```


# Custom models

The package is designed to allow users to construct their own models.

Typical steps are:

1. Prepare environmental driver data.
2. Define one or more life stages.
3. Specify development and variability functions.
4. Write a custom timestep function.
5. Run simulations with `run_simulation()`.

Because development models, variability models and life-stage definitions are modular, they can be combined in many different ways.


# Further information

See the help pages for:

```r
?new_lifestage
?new_chilled_lifestage
?run_simulation
?prepare_drivers
?linear_development
?cum_normal_variability
```

for full details of the modelling framework.
