---
title: "Introduction to phenologysim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phenologysim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)

set.seed(1234)

```

# Introduction

`phenologysim` is an R package for modelling phenology and population dynamics in insects and other organisms whose development depends on environmental conditions.

The package is available on GitHub at https://github.com/John-Kean/phenologysim and can be installed by calling `remotes::install_github("John-Kean/phenologysim")`.

```{r setup}
library(phenologysim)

```

The package provides:

- A structured framework for modelling life stages
- Temperature-dependent development models
- Calendar-based development
- Daylength-dependent life stages (a form of diapause)
- Chilling-dependent development (another form of diapause)
- Specific variability in developmental requirements
- Cohort-based population simulation with demographic stochasticity
- Visualisation tools
- Worked examples

Although originally developed for insect phenology modelling, the framework can be applied to any organism where progression through life stages depends on temperature, time, chilling accumulation or photoperiod.


# Core concepts

All functions and objects in `phenologysim` can be integrated in whatever way the author needs, but it encourages a structured approach to phenology simulation consisting of four main components:

1. **Environmental drivers** such as temperature and daylength.
2. **Life stages** such as eggs, larvae, pupae and adults.
3. **A timestep function** describing transitions between life stages.
4. **A simulation engine** that applies the timestep repeatedly through time.

The package includes helper functions for each of these components.


# Environmental drivers

Environmental conditions ("drivers") are provided as a data frame containing columns:

- *date*, in the form "YYYY-MM-DD"
- *dayfraction*, being the fraction of the day
- *degreesC*, the temperature in °C
- *daylength*, as hours of light
- *daytrend*, being 1 if daylength is increasing, or -1 if it is decreasing

Rows must be sorted in sequential order, and the dayfractions for all rows with the same date should add to 1.

If daily min/max/mean temperatures are provided, these can be converted to the required format using `prepare_drivers()`. Similarly, the column for daytrend can be added automatically using the same function.

```{r create-drivers}
# Create fake data with a seasonal cycle
cos <- 5 * cos(seq(0, 2 * pi, length.out = 365))
drivers_raw <- data.frame(
  date = seq.Date(as.Date("2020-07-01"), by = "day", length.out = 365),
  Tmin  = rnorm(365, mean = 10 - cos, sd = 1),
  Tmax  = rnorm(365, mean = 20 - cos, sd = 1),
  Tmean = rnorm(365, mean = 15 - cos, sd = 1)
)

# Format these data for use
drivers <- prepare_drivers(drivers_raw, latitude = -37.44)

head(drivers)

```

Correctly formatted drivers can be plotted to show variation in both temperature and daylength...

```{r plot-drivers}
# Plot the first week of data in detail
plot_drivers(drivers[1:32, ])

```
...or summarised to daily min, max and mean values.

```{r plot-daily-drivers}
# Plot all data as daily summaries
plot_daily_drivers(drivers)

```

# Life stages

Life stages are represented as S3 objects.

A typical life stage specifies:

- a development model
- development parameters
- a variability model
- variability parameters

For example:

```{r demo-create-larvae}
larvae <- new_lifestage(
  name = "Larvae",
  devFunction = linear_development,
  devParameters = list(
    b = 10,   # base temperature required for development (°C)
    q = 200   # day degrees required to complete the life stage (°d)
  ),
  varFunction = cum_normal_variability,
  varParameters = list(
    sd = 0.1  # standard deviation in relative developmental requirement (mean = 1)
  )
)

larvae

```

A life stage may contain multiple cohorts of different ages and developmental states.

To add a cohort, use `add_cohort()`:

```{r demo-add-cohort}
# Add a cohort of 1200 individuals
larvae <- add_cohort(larvae, 1200)

# Add another cohort of 34 individuals who have completed 50% of their development
larvae <- add_cohort(larvae, 34, cumDevelopment = 0.5)

larvae

```

To see the total number of individuals in that life stage use `total()`:

```{r demo-total}
total(larvae)

```

To implement mortality or survival, use `survive()`. For a one-off mortality event just specify the survival proportion:

```{r demo-survive}
larvae <- survive(larvae, 0.9)  # Implement 10% mortality

total(larvae)

```

Note that binomial rounding is used to maintain integer population sizes and introduce demographic stochasticity. Therefore, `survive()` and similar functions will not always give the same result.

```{r demo-stochasticity}
total(survive(larvae, 0.9))
total(survive(larvae, 0.9))
total(survive(larvae, 0.9))

```




## Development models

Many development models are available.

For example, a simple linear degree-day model:

```{r fig-linear-development}
curve(
  linear_development(x, b = 10, q = 200),
  from = 0,
  to = 35,
  xlab = "Temperature (°C)",
  ylab = "Development rate (/d)"
)

```

More complex nonlinear models are also available, including:

- beta_development()
- briere1_development()
- briere2_development()
- gaussian_development()
- lactin1_development()
- lactin2_development()
- logan1_development()
- logan2_development()
- sigmoid_development()


## Variability models

Development is rarely identical among individuals.

Variability functions convert accumulated development into the proportion of a population that has completed a stage.

For example:

```{r fig-cum-normal-variability}
curve(
  cum_normal_variability(x, sd = 0.1),
  from = 0.5,
  to = 1.5,
  xlab = "Relative development",
  ylab = "Proportion of population completed"
)

```

Alternative variability functions include:

- logistic_variability()
- regniere1_variability()
- regniere2_variability()
- step_variability()
- weibull_variability()


A whole set of life stages can be set up as a named list.

```{r example-life-stages}
example_lifestages <- list(

  eggs = new_chilled_lifestage(
    name = "Eggs",
    devFunction = linear_development,
    devParameters = list(b = 7.4, q = 428),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1),
    chillFunction = linear_chill,
    chillParameters = list(T_chill = 10),
    chillResponse = function(cumChill) { 1 - 0.46 * exp(-0.0005 * cumChill) }
  ),

  larvae = new_lifestage(
    name = "Larvae",
    devFunction = briere2_development,
    devParameters = list(
      T0 = 12,
      Tu = 33,
      a = 4.7e-5,
      b = 4.2
    ),
    varFunction = weibull_variability,
    varParameters = list(a = 0.8, c = 3)
  ),

  pupae = new_lifestage(
    name = "Pupae",
    devFunction = linear_development,
    devParameters = list(b = 10, q = 200),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1)
  ),

  adults = new_lifestage(
    name = "Adults",
    devFunction = calendar_development,
    devParameters = list(Q = 14),
    varFunction = cum_normal_variability,
    varParameters = list(sd = 0.1)
  )
)

```


# Population processes

Authors can 

```{r example-timestep}
example_timestep <- function(
  population,
  drivers,
  survival = c(
    eggs   = 0.999,
    larvae = 0.99,
    pupae  = 0.99,
    adults = 0.9
  ),
  fecundity = 0.2) {

  p <- population
  d <- drivers
  f <- drivers$dayfraction

  # Survival and development
  p$eggs   <- p$eggs   |> survive(survival["eggs"],   f) |> develop(d)
  p$larvae <- p$larvae |> survive(survival["larvae"], f) |> develop(d)
  p$pupae  <- p$pupae  |> survive(survival["pupae"],  f) |> develop(d)
  p$adults <- p$adults |> survive(survival["adults"], f) |> develop(d)

  # Stages mature
  p$eggs   <- p$eggs   |> add_cohort(rpois(1, total(p$adults) * fecundity))
  p$larvae <- p$larvae |> add_cohort(maturing(p$eggs))
  p$pupae  <- p$pupae  |> add_cohort(maturing(p$larvae))
  p$adults <- p$adults |> add_cohort(maturing(p$pupae))

  p
}


```


# Generic simulation engine

The package offers an optional generic simulation engine, `run_simulation()`. The user supplies:

- a population definition
- initial population sizes
- a timestep function
- environmental drivers

The bundled example model implements a complete insect life cycle.


# Running the example simulation

```{r run-simulation}
set.seed(123)

pop <- example_lifestages
pop$eggs <- add_cohort(pop$eggs, 1000)

system.time({
results <- run_simulation(
  population = pop,
  timestep = example_timestep,
  drivers = drivers
)
})

head(results)

```

The output contains both environmental variables and life-stage abundances through time.

## Visualising population dynamics

The trajectory of each life stage can be visualised using:

```{r fig-area-plot}
plot_population_trajectory(
  results[, c("date", "eggs", "larvae", "pupae", "adults")]
)

```

This shows the seasonal dynamics generated by the example model.

A line plot may also be used:

```{r fig-line-plot}
plot_population_trajectory(
  results[, c("date", "eggs", "larvae", "pupae", "adults")],
  type = "line"
)

```


# Custom models

The package is designed to allow users to construct their own models.

Typical steps are:

1. Prepare environmental driver data.
2. Define one or more life stages.
3. Specify development and variability functions.
4. Write a custom timestep function.
5. Run simulations with `run_simulation()`.

Because development models, variability models and life-stage definitions are modular, they can be combined in many different ways.


# Further information

See the help pages for:

```r
?new_lifestage
?new_chilled_lifestage
?run_simulation
?prepare_drivers
?linear_development
?cum_normal_variability
```

for full details of the modelling framework.
