---
title: "Introduction to phenologysim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phenologysim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)

set.seed(1234)

```

`phenologysim` is an R package for modelling phenology and population dynamics in insects and other organisms whose development depends on environmental conditions.

The package is available on GitHub at https://github.com/John-Kean/phenologysim and can be installed by calling `remotes::install_github("John-Kean/phenologysim")`.

```{r setup}
library(phenologysim)

```

The package provides:

- A structured framework for modelling life stages
- Temperature-dependent development models
- Calendar-based development
- Daylength-dependent life stages (a form of diapause)
- Chilling-dependent development (another form of diapause)
- Specific variability in developmental requirements
- Cohort-based population simulation with demographic stochasticity
- Visualisation tools
- Worked examples

Although originally developed for insect phenology modelling, the framework can be applied to any organism where progression through life stages depends on temperature, time, chilling accumulation or photoperiod.


# Core concepts

All functions and objects in `phenologysim` can be integrated in whatever way the author needs, but it encourages a structured approach to phenology simulation consisting of four main components:

1. **Environmental drivers** such as temperature and daylength.
2. **Life stages** such as eggs, larvae, pupae and adults.
3. **A timestep function** describing transitions between life stages.
4. **A simulation engine** that applies the timestep repeatedly through time.

The package includes helper functions for each of these components.


# Environmental drivers

Environmental conditions ("drivers") are provided as a data frame containing columns:

- *date*, in the form "YYYY-MM-DD"
- *dayfraction*, being the fraction of the day
- *degreesC*, the temperature in °C
- *daylength*, as hours of light
- *daytrend*, being 1 if daylength is increasing, or -1 if it is decreasing

Rows must be sorted in sequential order, and the dayfractions for all rows with the same date should add to 1.

If daily min/max/mean temperatures are provided, these can be converted to the required format using `prepare_drivers()`. Similarly, the column for daytrend can be added automatically using the same function.

```{r create-drivers}
# Create fake data with a seasonal cycle
cos <- 5 * cos(seq(0, 2 * pi, length.out = 365))
drivers_raw <- data.frame(
  date = seq.Date(as.Date("2020-07-01"), by = "day", length.out = 365),
  Tmin  = rnorm(365, mean = 10 - cos, sd = 1),
  Tmax  = rnorm(365, mean = 20 - cos, sd = 1),
  Tmean = rnorm(365, mean = 15 - cos, sd = 1)
)

# Format these data for use
drivers <- prepare_drivers(drivers_raw, latitude = -37.44)

head(drivers)

```

Correctly formatted drivers can be plotted to show variation in both temperature and daylength...

```{r plot-drivers}
# Plot the first week of data in detail
plot_drivers(drivers[1:32, ])

```
...or summarised to daily min, max and mean values.

```{r plot-daily-drivers}
# Plot all data as daily summaries
plot_daily_drivers(drivers)

```

# Life stages

Life stages are represented as S3 objects.

A typical life stage specifies:

- a development model
- development parameters
- a variability model
- variability parameters

For example:

```{r demo-new-lifestage}
instar1 <- new_lifestage(
  name = "First instar larvae",
  devFunction = linear_development,
  devParameters = list(
    b = 10,   # base temperature required for development (°C)
    q = 100   # day degrees required to complete the life stage (°d)
  ),
  varFunction = cum_normal_variability,
  varParameters = list(
    sd = 0.1  # standard deviation in relative developmental requirement (mean = 1)
  )
)

instar1

```

A life stage may contain multiple cohorts of different ages and developmental states.

To add a cohort, use `add_cohort()`:

```{r demo-add-cohort}
# Add a cohort of 1200 individuals
instar1 <- add_cohort(instar1, 1200)

# Add another cohort of 34 individuals who have completed 50% of their development
instar1 <- add_cohort(instar1, 34, cumDevelopment = 0.5)

instar1

```

To see the total number of individuals in that life stage use `total()`:

```{r demo-total}
total(instar1)

```

## Survival and mortality

To implement mortality or survival, use `survive()`. For a one-off mortality event just specify the survival proportion:

```{r demo-survive}
instar1 <- survive(instar1, 0.9)  # Implement 10% mortality

total(instar1)

```

Note that binomial rounding is used to maintain integer population sizes and introduce demographic stochasticity. Therefore, `survive()` and similar functions will not always give the same result.

```{r demo-stochasticity}
total(survive(instar1, 0.9))
total(survive(instar1, 0.9))
total(survive(instar1, 0.9))
total(survive(instar1, 0.9))
total(survive(instar1, 0.9))

```

Using `survive()` in this way implements an instant mortality. Often, mortality should be a more gradual process. If survival is to be implemented as a gradual rate, then it needs to be scaled for the length of the timestep by passing **day_fraction** (between 0 and 1) to the function:

```{r demo-scaled-survival}
# Apply survival at a rate of 0.9 /day across a period of 0.25 days (6 hours)
instar1 <- survive(instar1, 0.9, day_fraction = 0.25)

total(instar1)

```

It is also possible to remove all individuals from a life stage using `kill_all()`:

```{r demo-kill-all}
instar1 <- kill_all(instar1)

total(instar1)

```


## Development models

Many development models are available. For example, a simple linear degree-day model:

```{r fig-linear-development}
curve(
  linear_development(x, b = 10, q = 200),
  from = 0,
  to = 35,
  xlab = "Temperature (°C)",
  ylab = "Development rate (/d)"
)

```

or a non-linear Lactin model:

```{r fig-lactin2-development}
curve(
  lactin2_development(x, rho = 0.15, Tu = 33, delta = 6.4, lambda = -0.3),
  from = 0,
  to = 35,
  xlab = "Temperature (°C)",
  ylab = "Development rate (/d)"
)

```

Many other development rate models are available, including:

- beta_development()
- briere1_development()
- briere2_development()
- gaussian_development()
- lactin1_development()
- lactin2_development()
- logan1_development()
- logan2_development()
- sigmoid_development()

Variability in development is handled explicitly, using functions which relate accumulated development to the proportion of a population that has completed a stage.

For example:

```{r fig-cum-normal-variability}
curve(
  cum_normal_variability(x, sd = 0.1),
  from = 0.5,
  to = 1.5,
  xlab = "Relative development",
  ylab = "Proportion of population completed"
)

```

Other variability functions include:

- logistic_variability()
- regniere1_variability()
- regniere2_variability()
- step_variability()
- weibull_variability()

Life stage development is implemented by calling `develop()` for each life stage exactly once per timestep, passing the environmental drivers for that step. This calculates how many individuals should mature from that timestep and removes them from the cohort.

```{r demo-development}
instar1 <- add_cohort(instar1, 1000, cumDevelopment = 0.95)

instar1 <- develop(instar1, drivers[180, ])

instar1

```

The maturing individuals are then available through `maturing()` to be added to the next life stage using `add_cohort()`.

```{r demo-maturation}
instar2 <- new_lifestage(
  name = "Second instar larvae",
  devFunction = linear_development,
  devParameters = list(
    b = 10,  # base temperature required for development (°C)
    q = 50   # day degrees required to complete the life stage (°d)
  ),
  varFunction = cum_normal_variability,
  varParameters = list(
    sd = 0.1  # standard deviation in relative developmental requirement (mean = 1)
  )
)

new_instar2 <- maturing(instar1)

instar2 <- add_cohort(instar2, new_instar2)

instar2

```


# Simulating populations

It is often useful to set up a whole series of life stages at once, saving them as a named list.

```{r example-life-stages}
example_lifestages <- list(

  eggs = new_chilled_lifestage(
    name = "Eggs",
    devFunction = linear_development,
    devParameters = list(
      b = 7.4,
      q = 428
    ),
    varFunction = cum_normal_variability,
    varParameters = list(
      sd = 0.1
    ),
    chillFunction = linear_chill,
    chillParameters = list(
      T_chill = 10
    ),
    chillResponse = function(cumChill) { 1 - 0.46 * exp(-0.0005 * cumChill) }
  ),

  larvae = new_lifestage(
    name = "Larvae",
    devFunction = briere2_development,
    devParameters = list(
      T0 = 12,
      Tu = 33,
      a = 4.7e-5,
      b = 4.2
    ),
    varFunction = weibull_variability,
    varParameters = list(
      a = 0.8, 
      c = 3
    )
  ),

  pupae = new_lifestage(
    name = "Pupae",
    devFunction = linear_development,
    devParameters = list(
      b = 10, 
      q = 250
    ),
    varFunction = cum_normal_variability,
    varParameters = list(
      sd = 0.1
    )
  ),

  adults = new_lifestage(
    name = "Adults",
    devFunction = calendar_development,
    devParameters = list(
      Q = 14
    ),
    varFunction = cum_normal_variability,
    varParameters = list(
      sd = 0.1
    )
  )
)

```

Authors can then define a timestep function that captures the population processes

```{r example-timestep}
example_timestep <- function(
  population,
  drivers,
  survival = c(
    eggs   = 0.999,
    larvae = 0.99,
    pupae  = 0.99,
    adults = 0.9
  ),
  fecundity = 0.2) {

  p <- population
  d <- drivers
  f <- drivers$dayfraction

  # Survival and development
  p$eggs   <- p$eggs   |> survive(survival["eggs"],   f) |> develop(d)
  p$larvae <- p$larvae |> survive(survival["larvae"], f) |> develop(d)
  p$pupae  <- p$pupae  |> survive(survival["pupae"],  f) |> develop(d)
  p$adults <- p$adults |> survive(survival["adults"], f) |> develop(d)

  # Stages mature
  p$eggs   <- p$eggs   |> add_cohort(rpois(1, total(p$adults) * fecundity))
  p$larvae <- p$larvae |> add_cohort(maturing(p$eggs))
  p$pupae  <- p$pupae  |> add_cohort(maturing(p$larvae))
  p$adults <- p$adults |> add_cohort(maturing(p$pupae))

  p
}


```

Then the package offers a generic simulation engine, `run_simulation()`. The user supplies:

- the population as a named list of life stages with their initial population sizes
- a timestep function
- the environmental drivers

```{r demo-simulation}
set.seed(123)

pop <- example_lifestages
pop$eggs <- add_cohort(pop$eggs, 1000)

results <- run_simulation(
  population = pop,
  timestep = example_timestep,
  drivers = drivers
)

head(results)

```

The output contains both environmental variables and life stage abundances through time.


## Visualising population dynamics

The trajectory of each life stage can be visualised as an area plot using `plot_population_trajectory()`

```{r fig-area-plot}
plot_population_trajectory(
  results[, c("date", "eggs", "larvae", "pupae", "adults")]
)

```

Alternatively, a line plot may be generated.

```{r fig-line-plot}
plot_population_trajectory(
  results[, c("date", "eggs", "larvae", "pupae", "adults")],
  type = "line"
)

```


# Custom models

The package is designed to allow users to construct their own models.

Typical steps are:

1. Prepare environmental driver data.
2. Define one or more life stages.
3. Specify development and variability functions.
4. Write a custom timestep function.
5. Run simulations with `run_simulation()`.

Because development models, variability models and life-stage definitions are modular, they can be combined in many different ways.


# Further information

See the help pages for:

```r
?new_lifestage
?new_chilled_lifestage
?run_simulation
?prepare_drivers
?linear_development
?cum_normal_variability
```

for full details of the modelling framework.
